---
title: "Analysis Template"
output: html_document
date: '2022-05-30'
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup for text, include=FALSE}
library(dplyr)
library(Hmisc)
library(english)
library(stringr)
library(ggplot2)

################################################################################
# Input datasets from matching app
# Parameters - Match Parameters CSV
# Matches - Matched Data CSV
################################################################################
parameters <- read.csv("C:\\Matching Demo\\parameters.csv")


matches <- read.csv("C:\\Matching Demo\\matched_data.csv")



#set up dataset to be used with variable names and descriptions
description_table <- select(parameters, VALUE, description, PARAMETER) %>%
  subset(PARAMETER %in% c("Treatment Variable", "Matching Variable", "Outcome Variable"))

description_table$description <- capitalize(description_table$description) 

#Put all the variable names in the first set of quotes, and say whether the variable
#is considered "Demographics" or "Academic"
# parameters <- parameters %>%
#   mutate(var_type = case_when(VALUE == "URM_INDICATOR" ~ "Demographics",
#                               VALUE == "FIRST_GENERATION" ~ "Demographics",
#                               VALUE == "PELL_ELIGIBILITY" ~ "Demographics",
#                               VALUE == "RESIDENCY" ~ "Demographics",
#                               VALUE == "GENDER" ~ "Demographics",
#                               VALUE == "COHORT_TERM_CD" ~ "Academic",
#                               VALUE == "HIGH_SCHOOL_GPA" ~ "Academic",
#                               VALUE == "YEAR_START_STEM_RETN_STATUS" ~ "Academic",
#                               VALUE == "YEAR_START_MAJOR" ~ "Academic",
#                               VALUE == "ADMIT_STATUS" ~ "Academic",
#                               VALUE == "NEW_STUDENT_CATEGORY" ~ "Academic",
#                               VALUE == "CRS_LTTR_GRD" ~ "Academic",
#                               VALUE == "GPAO" ~ "Academic"))

# ################################################################################
# # Set up list of demographic and academic variables
# ################################################################################
# #This is taking all the variables that were indicated to be demographics and putting
# #them into a list, separated by commas and with "and" right before the last item
# #in the list
# demographics <- subset(parameters, parameters$var_type == "Demographics") %>%
#   select(description)
# demographics_list <- paste0(demographics$description, collapse=", ")
# demographics_list <- sub(",([^,]*)$", ", and \\1", demographics_list)
# 
# #This is taking all the variables that were indicated to be academics and putting
# #them into a list, separated by commas and with "and" right before the last item
# #in the list
# academics <- subset(parameters, parameters$var_type == "Academic") %>%
#   select(description)
# academics_list <- paste0(academics$description, collapse=", ")
# academics_list <- sub(",([^,]*)$", ", and \\1", academics_list)



################################################################################
#Set up list of outcome variables
################################################################################
outcome_variables <- subset(parameters, parameters$PARAMETER == "Outcome Variable") %>%
  select(description)

if (nrow(outcome_variables) == 1) {
  outcome_variables_list = outcome_variables$description
} else{
  outcome_variables$description_comma <- paste(outcome_variables$description, ", ", sep="")
  outcome_variables[nrow(outcome_variables),2] <- paste("and ", outcome_variables[nrow(outcome_variables),2], sep="")
  outcome_variables[nrow(outcome_variables),2] <- substr(outcome_variables[nrow(outcome_variables),2], 1,
                                                       nchar(outcome_variables[nrow(outcome_variables),2])-2)
  outcome_variables_list <- paste0(outcome_variables$description_comma, collapse="")
}

################################################################################
#Set up the ratio used as a word instead of a number, and get the type of matching
#algorithm used
################################################################################
#Grab the Matching Ratio from the parameter dataset
ratio <- parameters[parameters$PARAMETER == "Matching Ratio",2] 

#Turn the number into a word
ratio_text = as.character(english(as.numeric(ratio)))
ratio_text = capitalize(ratio_text)

#Grab the Matching Algorithm from the parameter dataset
algorithm <- parameters[parameters$PARAMETER == "Matching Algorithm", 2]

if (algorithm == "optimal") {
  algorithm_sentence = "using a weighted, optimal pair matching algorithm which minimized the global distance in characteristics between the treatment and control."
} else {
  algorithm_sentence = "using a greedy, nearest neighbors matching algorithm which matched each treatment to its most similar available control unit in a sequential manner."
}

#Grab the Matching Variables from teh parameter dataset
var_type_list<-parameters%>%
  dplyr::filter(!is.na(var_type))%>%
  dplyr::select(var_type)%>%
  distinct()
var_type_list<-var_type_list$var_type

var_type_list_element<-NULL
for(i in seq_along(var_type_list)){
var_type_list_element_i<-parameters%>%
    dplyr::filter(var_type==var_type_list[i])%>%
    dplyr::select(description)%>%
    distinct()
var_type_list_element_i<-paste(var_type_list_element_i$description,collapse=", ")
var_type_list_element<-c(var_type_list_element,var_type_list_element_i)
}

str_sub(paste(paste0(var_type_list,' factors ','(',var_type_list_element,') and'),collapse=" "),end=-5)

if(length(var_type_list)>0){
  matching_variable_sentence<-paste0(str_sub(paste(paste0(var_type_list,' factors ','(',var_type_list_element,') and'),collapse=" "),end=-5),'.')
}
if(length(var_type_list)==0){
  matching_variable_sentence<-'variety of factors.'
}
################################################################################
#Create dataset to make table with weights
################################################################################
weights <- subset(parameters, parameters$PARAMETER == "Matching Variable") %>%
  select(description, WEIGHT_OR_LEVEL) %>%
  arrange(desc(as.numeric(WEIGHT_OR_LEVEL)))

weights$description <- capitalize(weights$description)

################################################################################
#Sort matching variables for descriptive statistics
#Character variables are put into Tables
#Non-character variables are put into Figures
################################################################################
#Sort matching variables into with "Table" or "Figure" based on whether they are
#a character or non-character
parameters$table_figure <- ifelse(parameters$PARAMETER == "Matching Variable" & 
                       parameters$DATA_TYPE == "character", "Table", 
                       ifelse(parameters$PARAMETER == "Matching Variable" & 
                       parameters$DATA_TYPE != "character", "Figure", ""))

#Put the table variables into 1 dataset
table_variables <- subset(parameters, parameters$table_figure == "Table") %>%
  select(VALUE, description)

#Capitalize the first letter for every variable
table_variables$description <- capitalize(table_variables$description)

#Put the figure variables into 1 dataset
figure_variables <- subset(parameters, parameters$table_figure == "Figure") %>%
  select(VALUE, description)

#Capitalize the first letter for every variable
figure_variables$description <- capitalize(figure_variables$description)

################################################################################
#Set the treatment variable to the "TREATMENT_INDICATOR_LABEL" 
#This will be used in the regressions
#Also pull out what the treatment group is
################################################################################
treatment_variable_value <- "TREATMENT_INDICATOR_LABEL"

treatment <- subset(parameters, parameters$PARAMETER == "Treatment Variable") %>%
  select(VALUE, description, WEIGHT_OR_LEVEL)

treatment_value <- treatment[1,1]
treatment_description <- treatment[1,2]
treatment_condition <- treatment[1,3]

################################################################################
#Get the total sample size, and the sample size for the treatment and control groups
################################################################################
#Get the total sample size
sample_size <- nrow(matches)

#Get the treatment sample size
treatment_sample_size <- nrow(subset(matches, matches$TREATMENT_INDICATOR == 1))

#Get the control sample size
control_sample_size <- nrow(subset(matches, matches$TREATMENT_INDICATOR == 0))

################################################################################
#Sort the outcomes variables by whether they are a character of a non-character
################################################################################
#Create dataset for character outcome variables
outcome_variables_character <- subset(parameters, 
                                      parameters$PARAMETER == "Outcome Variable" &
                                        parameters$DATA_TYPE == "character") %>%
  select(VALUE, description)

#Create dataset for non-character outcome variables
outcome_variables_numeric <- subset(parameters, 
                                      parameters$PARAMETER == "Outcome Variable" &
                                        parameters$DATA_TYPE != "character") %>%
  select(VALUE, description)

################################################################################
#List of statistical tests
################################################################################
if (nrow(outcome_variables_character) !=  0) {
  outcome_tests_character = paste("chi-square tests^[A chi-squared test is used to examine whether the outcome of interest is independent of treatment status. The assumptions are (1) the categories are mutually exclusive, (2) the treatment and control group are independent, (3) the outcome of interest must be a categorical variable, and (4) the expected value of each cell must be 5 or greater. A p-value of less than 10% means that you reject the null hypothesis. This means there is a difference in the outcome variable of interest between the treatment and control group.]", "logistic regressions^[A logistic regression models the probability of an event taking place. Here we are testing whether treatment status affects the odds of the outcome of interest. The assumptions are (1) the outcome of interest must be binary, (2) the observations are independent, (3) no multicollinearity, (4) no extreme outliers, (5) there is a linear relationship between the outcome of interest and treatment status, and (6) the sample size is sufficiently large. A p-value of less than 10% means there is a statistically significant relationship between the outcome of interest and treatment status. To treatment status coefficient can be interpreted as e raised to the coefficient is the difference in odds of the outcome of interest between the treatment and control group.]", sep=", ")
} else {
  outcome_tests_character = ""
}

if (nrow(outcome_variables_numeric) != 0) {
  outcome_tests_numeric = paste("t-tests^[A t-test is used to to test whether the mean of the outcome of interest is the same between the treatment and control group. The assumptions are (1) the outcome variable must be continous or ordinal, (2) the data is collected from a simple random sample of the population, (3) data follows a normal distribution, (4) a reasonably large sample size is used, and (5) there is an equal variance between the two populations. A p-value of less than 10% means you can reject the null hypothesis. This means there is statistically significant evidence that the means are different between the treatment and control groups.]", "Wilcoxom rank sum tests^[The Wilcoxom rank sum test is a non-parametric version of the two-sample t-test. This means that, unlike the t-test, the population does not need to be normally distributed to use this test. This is testing to see if the two samples have equal medians. The assumptions are (1) the treatment and control group are independent of each other, and (2) the populations have equal variance. A p-value of less than 10% means there is statistically significant evidence that the medians for the variable of interest differ between the treatment and control group.]", "linear regressions^[Linear regressions are used to examine the relationship between the outcome of interest and treatment status. The assumptions are (1) there is a linear relationship between the outcome variable and treatment status, (2) the residuals are normally distributed, (3) the variance of the residuals are constant across treatment status, and (4) no autocorrelation of errors - the error term of one observation cannot be used to predict the error term of another observation]", sep=", ")
} else {
  outcome_tests_numeric = ""
}

permutation_test = "and permutation tests^[A permutation test is a nonparametric test used to estimate the population distribution and determine how rare our observed values are relative to the population. The only assumption is that the observations are exchangeable between the treatment and control groups. A p-value of less than 10% means the probability of obtaining the initial difference in the means of the outcome variable between the treatment and control groups occurs with a low probability, meaning there is statstically signicant evidence that the mean of the outcome variable for the treatment group is not equal to the mean of the outcome variable for the control group.]"

if (outcome_tests_character != "" & outcome_tests_numeric != "") {
  outcome_tests_list <- paste(outcome_tests_character, outcome_tests_numeric, permutation_test, sep=", ")
} else if (outcome_tests_character == "" & outcome_tests_numeric != "") {
  outcome_tests_list <- paste(outcome_tests_numeric, permutation_test, sep=", ")
} else {
  outcome_tests_list <- paste(outcome_tests_character, permutation_test, sep=", ")
}
```

## Outcomes

**Description**

A matching analysis was performed to create a treatment and control group. The treatment and control groups were determined by `r treatment_description` (`r treatment_value`) where the treatment group were students with `r treatment_value` equal to `r treatment_condition`. The students were matched on `r matching_variable_sentence` `r ratio_text`-to-one matching was performed `r algorithm_sentence`^[The foundation of any matched analysis is a distance measure, which quantifies how similar two units of analysis (i.e. students) are to each other. The smaller the distance between the two units, the more similar they are and the greater likelihood that the two units should be paired together in a matched analysis. This matched analysis used a distance measure that is a variation of the Gower distance formula (<https://www.researchgate.net/publication/348320805_Distances_with_mixed_type_variables_some_modified_Gower's_coefficients>). The distance measure in this analysis represented the average distance between any pair of treatment and control units, with each matching variable being weighted using the weights represented in the “Matching Weights” table. For numeric variables, the distance is calculated as the absolute value of the difference between the two units on the variable. Prior to calculating that difference, numerical variables are normalized between 0 and 1 by dividing each value by the maximum value of the variable in the dataset. For categorical variables, the distance is set to 0 if the two units match exactly on the level of the variable and the distance is set to 1 if the two units do not match exactly on the variable. If either the ith treatment or jth control unit have a missing value on the variable, the distance is set to 1 to reflect the uncertainty of that variable in the matching process for the potential matched pair.] The matching characteristics/variables were weighted prior to running the matching algorithm. The "Matching Weights" table shows the matching variables and their respective weights. The matching yielded a total sample size of `r sample_size` students for analysis, with `r treatment_sample_size` students in the treatment group and `r control_sample_size` students in the control group. After matching, the balance of the matching factors between the treatment and control was confirmed using chi-square tests and t-tests.


```{r intro tables, echo=FALSE, message=FALSE, warning=FALSE}
# This creates a table with the variable names and a desciption of the variable
library(knitr)
library(kableExtra)

kable(description_table, caption = "Variables and Descriptions", col.names = c("Variable", "Description", "Type"),
        row.names = FALSE) %>%
    kable_styling(full_width = FALSE, position = "center")

kable(weights, caption = "Matching Weights", col.names = c("Matching Variable", "Weight"))  %>%
    kable_styling(full_width = FALSE, position = "center")

```

After the treatment and control group is created contigency tables and box plots are generated to ensure the control and treatment group are similar for the matching variables.

### Comparison of Treatment and Control Groups for Matching Variables
```{r comparison, results = "asis", echo=FALSE, message=FALSE, warning=FALSE, fig.align = 'center'}
library(janitor)
library(knitr)
library(kableExtra)

#This loops through all the character variables and creates a contingency table
#for each
if (nrow(table_variables) > 0) {
  for (row in 1:nrow(table_variables)) {
    row_name = table_variables[row, 1]
    row_name_desc = table_variables[row, 2]
    table1 <- tabyl(matches, !!as.name(row_name), TREATMENT_INDICATOR_LABEL) %>%
      adorn_percentages("col") %>%
      adorn_pct_formatting(digits=1) %>%
      kable(caption = paste(row_name_desc, "Contingency Table", sep=" "), 
            col.names = c("", "Control", "Treatment")) %>%
      kable_styling(full_width = FALSE) %>%
      column_spec(1, width = "3in") %>%
      print()
  
  }
}

#This loops through all the non-character variables and creates a boxplot for each one
if (nrow(figure_variables) > 0) {
  for (row in 1:nrow(figure_variables)) {
    row_name = figure_variables[row, 1]
    row_name_desc = figure_variables[row, 2]
    print(ggplot(matches, aes(x=TREATMENT_INDICATOR_LABEL, y=!!as.name(row_name), 
                              fill=TREATMENT_INDICATOR_LABEL)) + xlab("") + 
            theme(legend.position = "none") + 
            ggtitle(paste(row_name_desc, "Box Plot", sep=" ")) +
            ylab(row_name_desc) + geom_boxplot())
  }
}
```

### Statistical Analysis of Outcomes

After the control group was created and matching balance was confirmed, the treatment group was statistically compared with the control group in terms of `r outcome_variables_list`. The methods used were `r outcome_tests_list`.

```{r analysis, results = "asis", echo=FALSE, message=FALSE, warning=FALSE, fig.align = 'center'}
library(dplyr)
library(kableExtra)
library(stargazer)

#This creates a function that takes a y and x value that represent columns in a 
#dataset. It then runs a logistic regression with those variables
logit <- function(y, x){
  y_col <- as.name(y)
  x_col <- as.name(x)
  f <- substitute(glm(y_col~x_col, data=matches, family=binomial))
  eval(f)
}

#This creates a function that takes a y and x value that represent columns in a 
#dataset. It then runs a linear regression with those variables
regression <- function(y, x){
  y_col <- as.name(y)
  x_col <- as.name(x)
  f <- substitute(lm(y_col~x_col, data=matches))
  eval(f)
}

#Permutation test
permutation.test <- function(treatment, outcome, n){
  distribution=c()
  result=0
  for (i in 1:n){
    distribution[i]=diff(by(outcome, sample(treatment, length(treatment), FALSE), mean, na.rm=TRUE))
  }
  result=sum(abs(distribution) >= abs(original))/(n)
  return(list(result, distribution))
}

#Loop through every outcome variable that is a character and (1) creates a header, 
#(2) creates and outputs a contigency tables, (3) runs a chi-squared, and (4) runs
#a logist regression
if (nrow(outcome_variables_character) > 0) {
  for (row in 1:nrow(outcome_variables_character)) {
     cat("  \n")
     cat("  \n")
     
     row_name = outcome_variables_character[row, 1]
     row_name_desc = capitalize(outcome_variables_character[row, 2])
     
     cat(paste0("**Outputs: ", row_name_desc, "**\n\n"))
     
     unique <- unique(matches[[row_name]])
     unique <- subset(unique, !is.na(unique))
     
     if (length(unique) <= 2) {
       eval <- select(matches, row_name, treatment_variable_value)
       eval_cont <- table(eval)
       
       chi <- chisq.test(eval_cont)
       
       eval_cont_table_expected <- chi$expected
       
       min <- min(eval_cont_table_expected)
       
       if (min < 5){
         kable(eval_cont, caption = "Contingency Table") %>%
           add_header_above(c("", setNames(2,treatment_variable_value))) %>%
           kable_styling(full_width = FALSE) %>%
           column_spec(1, width = "0.5in") %>%
           print()
         
         cat("Unable to perform a chi-square test because expected cell counts are too small")
       } else {
         t1 <- data.frame(stat = c("X-squared", "df", "p-value"),
                          value = c(chi$statistic, chi$parameter, chi$p.value))
       
         kable(eval_cont, caption = "Contingency Table") %>%
           add_header_above(c("", setNames(2,treatment_variable_value))) %>%
           kable_styling(full_width = FALSE, position = "float_left") %>%
           column_spec(1, width = "0.5in") %>%
           print()
    
         kable(t1, col.names = NULL, row.names = FALSE, caption = "Chi-Square Results") %>%
           add_header_above(c("", "Pearson's Chi-squared test" = 1)) %>%
           kable_styling(full_width = FALSE, position = "center") %>%
           column_spec(1, width = "1in") %>%
           print()
       }
       
       matches$y <- ifelse(matches[[row_name]]=="Y", 1, 0)
       
       matches[[row_name]] <- matches$y
    
       logit1 <- logit(row_name, treatment_variable_value)
       stargazer(logit1, type = "html", header=FALSE,
                 title="Logistic Regression",
                 star.char = c("*", "**", "***"),
                 star.cutoffs = c(.1, .05, .01),
                 notes = c("Note: * p<0.1; ** p<0.05; *** p<0.01"),
                 notes.append=FALSE)
       
       #Permutation
       original <- diff(tapply(matches[[row_name]], matches[[treatment_variable_value]], mean, na.rm=TRUE))
      
       test1 <- permutation.test(matches[[treatment_variable_value]], matches[[row_name]], 10000)
       
       hist(test1[[2]], breaks=50, col='grey', main="Permutation Distribution of Mean Treatment - Mean Control", las=1, xlab='', sub="The red line is the original difference we oberved.")
    abline(v=original, lwd=3, col="red") 
    
    p = sum(abs(unlist(test1)) > abs(original)) / 10000
    
    cat("  \n")
    cat("The probability of obtaining means that are not equal is", p)
  
    cat("  \n")
    cat("  \n")
     } else {
       cat("This document does not analyze nominal outcome variables with more than two levels.")
     }
  }
}

#Loop through every outcome variable that is a non-character and (1) creates a header, 
#(2) runs a t-test, and (3) runs a logistic regression
if (nrow(outcome_variables_numeric) > 0) {
  for (row in 1:nrow(outcome_variables_numeric)) {
    cat("  \n")
    cat("  \n")
  
    row_name = outcome_variables_numeric[row, 1]
    row_name_desc = capitalize(outcome_variables_numeric[row, 2])
     
    cat(paste0("**Outputs: ", row_name_desc, "**\n\n"))
  
    t_test <- t.test(matches[[row_name]]~matches[[treatment_variable_value]])
  
    t2 <- data.frame(stat = c("t", "df", "p-value"),
                     value = c(t_test$statistic, t_test$parameter, t_test$p.value))
  
    kable(t2, col.names = NULL, row.names = FALSE, caption = "t-test Results") %>%
      add_header_above(c("", "Welch Two Sample t-test" = 1)) %>%
      kable_styling(full_width = FALSE, position = "float_left") %>%
      column_spec(1, width = "1in") %>%
      print()
    
    wilcox_test <- wilcox.test(matches[[row_name]]~matches[[treatment_variable_value]], data=matches)
    
    w2 <- data.frame(stat = c("t", "p-value"),
                     value = c(wilcox_test$statistic, wilcox_test$p.value))
    
    kable(w2, col.names = NULL, row.names = FALSE, caption = "Wilcoxon Rank Sum Results") %>%
      add_header_above(c("", "Wilcoxon Rank Sum Test" = 1)) %>%
      kable_styling(full_width = FALSE, position = "center") %>%
      column_spec(1, width = "1in") %>%
      print()
  
    reg1 <- regression(row_name, treatment_variable_value)
    stargazer(reg1, type = "html", header=FALSE, align=F,
              title="Linear Regression",
              star.char = c("*", "**", "***"),
              star.cutoffs = c(.1, .05, .01),
              notes = c("Note: * p<0.1; ** p<0.05; *** p<0.01"),
              notes.append=FALSE) 
    
    #Permutaation
    original <- diff(tapply(matches[[row_name]], matches[[treatment_variable_value]], mean, na.rm=TRUE))
    
    test1 <- permutation.test(matches[[treatment_variable_value]], matches[[row_name]], 10000)
    
    hist(test1[[2]], breaks=50, col='grey', main="Permutation Distribution of Mean Treatment - Mean Control", las=1, xlab='', sub="The red line is the original difference we oberved.")
    abline(v=original, lwd=3, col="red") 
    
    p = sum(abs(unlist(test1)) > abs(original)) / 10000
    
    cat("  \n")
    cat("The probability of obtaining means that are not equal is", p)
  
    cat("  \n")
    cat("  \n")
  }
}
```
